Wigglyhedra

### 2024-06-07 draft/wigglyhedra.tex
Updated categorical constraints (lots of todos still).

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{assumption}[theorem]{Assumption}
\crefname{notation}{Notation}{Notations}
\crefname{problem}{Problem}{Problems}

@@

% operators
\DeclareMathOperator{\conv}{conv} % convex hull
\DeclareMathOperator{\vect}{vect} % linear span
\DeclareMathOperator{\cone}{cone} % cone hull
\DeclareMathOperator{\inv}{inv} % inversions
\DeclareMathOperator{\ninv}{ninv} % inversions
\DeclareMathOperator{\Ima}{Im} % image
\DeclareMathOperator{\Vol}{Vol} % (mixed) volume
\DeclareMathOperator{\Hom}{Hom} % hom-spaces

@@

See~\cite[Section 3]{hue.kho:01} for further details on the definition, and~\cite[Remark 6.6]{bap.deo.lic:20} for more discussion.
Let \(P_i\) denote the indecomposable projective \(A_m\)-module defined as \(P_i = A_m\cdot e_i\).
For any integer \(j\), let \(P_i \langle j \rangle\) denote its graded shift by \(j\).
Set \(\mathcal{C}_m\) to be the smallest triangulated subcategory of \(D^b(A_m\text{-mod})\) containing all the objects \(P_i\langle j \rangle\) (see, e.g.,~\cite[Section 2.3.3]{bap.deo.lic:22} for further discussion on the construction).

The category \(\mathcal{C}_m\) is \(k\)-linear and triangulated, where \([1]\) denotes the triangulated shift.
For \(X, Y \in \mathcal{C}_m\), let \(\Hom^{i,j}(X,Y)\) denote the space of morphisms in \(\mathcal{C}_m\) of homological degree \(i\) and internal degree \(j\).
In other words, we have
\[\Hom^{i,j}(X,Y) = \Hom(X,Y\langle i \rangle[j]).\]
Furthermore, for any integers \(a\) and \(b\), we have
\[\Hom^{i,j}(X\langle a \rangle[b], Y\langle a \rangle[b]) = \Hom^{i,j}(X,Y).\]

\begin{remark}\label{rem:generating-morphisms}
  The generating objects \(P_i\) (and hence their graded and homological shifts) satisfy the following:
  \begin{itemize}
  \item \(\Hom^{a,b}(P_i,P_i) = 0\) if \(b \neq 0\) and \(a \neq 0, 2\);
  \item \(\Hom^{a,0}(P_i,P_i) = k\) if \(a = 0, 2\);
  \item \(\Hom^{a,b}(P_i, P_{i\pm1}) = 0\) if \(b \neq 0\) and \(a \neq 1\);
  \item \(\Hom^{1,0}(P_i, P_{i\pm1}) = k\), and
  \item \(\Hom^{a,b}(P_i, P_j) = 0 \) otherwise.
  \end{itemize}
\end{remark}

The category \(\mathcal{C}_m\) has a standard bounded \(t\)-structure, whose heart consists of the \emph{linear complexes}.
\asilata{Say more about linear complexes.}

\subsection{The Khovanov--Seidel curve model}
In this section we briefly recall the procedure to recover an object of \(\mathcal{C}_{n+1}\) from a wiggly arc on \(n+2\) points on a line.
In fact, the procedure works on more general arcs as well --- namely, those that are not necessarily \(x\)-monotone.
All the results of this section are due to Khovanov and Seidel~\cite{kho.sei:02}.

@@

\begin{lemma}
Every wiggly arc corresponds to an object in the heart of the standard \(t\)-structure.  
\end{lemma}
\begin{proof}
  The procedure to build up the complex of projective modules from a curve explicitly describes the differentials.
  Each component of these differentials is a map from a shift of \(P_i\) to a shift of \(P_j\).
  It can be seen from~\cref{rem:generating-morphisms} that any component that maps from \(P_i\) to \(P_j\) for \(i \neq j\) keeps the complex linear; it is only when we have a looping map from \(P_i\) to \(P_i\langle 2 \rangle\) that we move out of the standard heart.
  \asilata{Now figure out what to say about \(t\)-structures, namely that wiggly arcs are exactly those whose associated complexes lie in the heart of the standard \(t\)-structure.}
\end{proof}

\subsection{Categorical conditions on compatibility of arcs}

\begin{proposition}
  Let \(X \in \mathcal{C}_{n+1}\) be a spherical object.
  Consider its cohomology filtration
  \[
    \begin{tikzcd}[column sep=0.3cm]
      0 = X_0 \arrow{rr} && X_1 \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{k-1} \arrow{rr} \arrow{dl} && X_k = X \arrow{dl}\\
      & H^1(X) \arrow[dashed]{ul}{+1} && H^2(X) \arrow[dashed]{ul}{+1} && \cdots \arrow[dashed]{ul} &&H^k(X) \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]
  Consider the set of objects \(\{H^j(X)\}\) taken without multiplicity, and let \(T\) be corresponding set of wiggly curves.
  We have the following results.
  \begin{enumerate}
  \item 
    \(T\) is a subset of a wiggly pseudotriangulation.
  \item \(T\) contains at most one of the two irrelevant arcs.
  \end{enumerate}
\end{proposition}

>>>>

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{defprop}[theorem]{Definition/Proposition}
\crefname{notation}{Notation}{Notations}
\crefname{problem}{Problem}{Problems}

@@

% operators
\DeclareMathOperator{\conv}{conv} % convex hull
\DeclareMathOperator{\vect}{vect} % linear span
\DeclareMathOperator{\cone}{cone} % cone hull
\DeclareMathOperator{\inv}{inv} % inversions
\DeclareMathOperator{\ninv}{ninv} % inversions
\DeclareMathOperator{\Ima}{Im} % image
\DeclareMathOperator{\Vol}{Vol} % (mixed) volume
\DeclareMathOperator{\Hom}{Hom} % hom-spaces
\DeclareMathOperator{\Ext}{Ext} % extensions

@@

\begin{remark}
  The definition of the zigzag algebra as used here originally appeared in~\cite[Section 3]{hue.kho:01}.
  This is an ``un-signed'' version of the zigzag algebra.
  It is technically better to use a ``signed'' version of the zigzag algebra, and the comparison between the two is discussed in~\cite[Remark 6.6]{bap.deo.lic:20}.
  As the two versions are isomorphic in our example, we ignore this technicality and use the easier definition.
\end{remark}

Let \(P_i\) denote the indecomposable projective \(A_m\)-module defined as \(P_i = A_m\cdot e_i\).
For any integer \(j\), let \(P_i \langle j \rangle\) denote its graded shift by \(j\).
Let \(\mathcal{C}_m\) be the smallest triangulated subcategory of \(D^b(A_m\text{-mod})\) containing all the objects \(P_i\langle j \rangle\) (see, e.g.,~\cite[Section 2.3.3]{bap.deo.lic:22} for further discussion on the construction).

The category \(\mathcal{C}_m\) is \(k\)-linear and triangulated, where \([1]\) denotes the triangulated shift.
For \(X, Y \in \mathcal{C}_m\), let \(\Hom^{g,h}(X,Y)\) denote the space of morphisms in \(\mathcal{C}_m\) of internal degree \(g\) and homological degree \(h\).
In other words, we have
\[\Hom^{g,h}(X,Y) = \Hom(X,Y\langle g \rangle[h]).\]
Furthermore, for any integers \(a\) and \(b\), we have
\[\Hom^{g,h}(X\langle g' \rangle[h'], Y\langle g' \rangle[h']) = \Hom^{g,h}(X,Y).\]

\begin{remark}\label{rem:generating-morphisms}
  It is easy to check that the generating objects \(P_i\) (and hence their graded and homological shifts) satisfy the following:
  \begin{itemize}
  \item \(\Hom^{g,h}(P_i,P_i) = 0\) if \(g \neq 0, 2\) and \(h \neq 0\);
  \item \(\Hom^{g,0}(P_i,P_i) = k\) if \(g = 0, 2\);
  \item \(\Hom^{g,h}(P_i, P_{i\pm1}) = 0\) if \(g \neq 1\) and \(h \neq 0\);
  \item \(\Hom^{1,0}(P_i, P_{i\pm1}) = k\), and
  \item \(\Hom^{g,h}(P_i, P_j) = 0 \) otherwise.
  \end{itemize}
\end{remark}

We now describe the full subcategory of \emph{linear complexes} of \(\mathcal{C}_m\).
It is convenient to think of an object \(P^{\bullet}\) of \(\mathcal{C}_m\)
as bigraded (projective) \(A_m\)-module equipped with a differential \(d \colon P^{\bullet} \to P^{\bullet}\).
More precisely, we can write
\[P^{\bullet} = \bigoplus M_i\langle g \rangle[h],\]
where \(g\) is the internal grading shift, and \(h\) is the homological grading shift, and 
each object \(M_i\) is a direct sum of copies of objects in \(\{P_1,\ldots, P_m\}\).

Recall that an object of the form \(M_i[h]\) is placed in homological degree \(-h\);
since the differential \(d\) raises the homological degree by one, it \emph{lowers} the homological shift by one.
That is, we have
\[d \colon M_i\langle g \rangle[h] \to \bigoplus M_{i'}\langle g' \rangle[h-1].\]
In other words, \(d\) maps \(P^h\) to \(P^{h-1}\).
\begin{definition}\label{def:linear-complex}
  Let \(P^{\bullet}\) be an object of \(\mathcal{C}_m\), described as a bigraded module
  \[P^{\bullet} = \bigoplus M_i\langle g \rangle[h]\]
  together with an internal differential \(d \colon P^{\bullet} \to P^{\bullet + 1}\).
  We say that \(P^{\bullet}\) has \emph{level \(\ell\)} for some \(\ell \in \mathbb{Z}\) if for each bigraded piece \(M_i \langle g \rangle[h]\) of \(P^{\bullet}\), we have \(g + h = \ell\).
  We say that \(P^{\bullet}\) is \emph{linear} if \(P^{\bullet}\) has level \(\ell = 0\).

\end{definition}
Let \(\mathcal{A}_m\) be the smallest full subcategory of \(\mathcal{C}_m\) that is closed under isomorphism, which contains all the linear complexes.
\begin{proposition}\label{prop:standard-heart}
  The category \(\mathcal{A}_m\) is abelian.
  Furthermore, it is the heart of a bounded \(t\)-structure on \(\mathcal{C}_m\), which we call the \emph{standard \(t\)-structure} on \(\mathcal{C}_m\).
\end{proposition}
\asilata{Find a reference for this and possibly a cleaner statement.}

Also recall the following statements for future use.
\begin{defprop}\label{defprop:cohomology}
  Let \((\mathcal{C}^{\le 0}, \mathcal{C}^{\geq 0})\) be a bounded \(t\)-structure on a triangulated category \(\mathcal{C}\) with heart \(\mathcal{A} = \mathcal{C}^{\leq 0} \cap \mathcal{C}^{\geq 0}\).
  For each integer \(i\), we have the \(i\)th truncation functors
  \[\tau^{\leq i} \colon \mathcal{C} \to \mathcal{C}^{\leq i}\]
  and
  \[\tau^{\geq i} \colon \mathcal{C} \to \mathcal{C}^{\geq i},\]
  which are left and right adjoint respectively\asilata{Check left vs right.} to the inclusion functors \(\mathcal{C}^{\leq 0} \to \mathcal{C}\) and \(\mathcal{C}^{\geq 0} \to \mathcal{C}\).
  The \(i\)th cohomology functor is defined to be the composition \(\tau^{\leq i} \circ \tau^{\geq i}\).
  It is equal to the composition \(\tau^{\geq i} \circ \tau^{\leq i}\), and maps an object of \(\mathcal{C}\) to an object of \(\mathcal{A}\).
\end{defprop}
In the presence of a bounded \(t\)-structure, every object of \(\mathcal{C}\) has a canonical filtration whose factors are shifts of its cohomology objects.
\begin{defprop}\label{defprop:cohomology-filtration}
  Let \(\mathcal{C}\) be any triangulated category, and \(\mathcal{A} \subset \mathcal{C}\) the heart of a bounded \(t\)-structure on \(\mathcal{C}\).
  Let \(X\) be a non-zero object of \(\mathcal{C}\), and let \(k\) and \(l\) be the maximum and minimum integers respectively for which \(H^k(X)\) and \(H^{l}(X)\) are non-zero.
  \asilata{Find references for cohomology and cohomology filtration.}
  Then \(X\) has a unique filtration of the form
  \[
    \begin{tikzcd}[column sep=0.1cm]
      0  \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k+1} \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} && \cdots \arrow[dashed]{ul} &&H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]  
\end{defprop}

\subsection{The Khovanov--Seidel curve model}
In this section we briefly recall how to recover an object of \(\mathcal{C}_{n+1}\) from a wiggly arc on \(n+2\) points on a line.
In fact, the procedure works on more general arcs as well --- namely, those that are not necessarily \(x\)-monotone.
All the results of this section are due to Khovanov and Seidel~\cite{kho.sei:02}.

@@

\begin{lemma}
  Every wiggly arc corresponds to an object in the heart \(\mathcal{A}_m\) of the standard \(t\)-structure on \(\mathcal{C}_m\).
\end{lemma}
\begin{proof}
  Let \(P^{\bullet}\) be the complex built from an arbitrary (not necessarily wiggly) arc.
  The procedure to build up the complex of projective modules from a curve explicitly describes the differentials.

  Each component of these differentials is a map of the form
  \[P_i\langle g \rangle[h] \to P_j\langle g' \rangle[h-1].\]
  First consider a component of the differential from some \(P_i\langle g \rangle\) to some \(P_j\langle g' \rangle\) for \(i \neq j\).
  We see from~\cref{rem:generating-morphisms} that any such map must take \(P_i\langle g \rangle\) to \(P_j \langle g+1 \rangle\).
  In other words, the differentials has the form
  \[P_i\langle g \rangle[h] \to P_j\langle g+1 \rangle[h-1]\]
  if \(i \neq j\).
  Since \(P_i\langle g \rangle[h]\) and \(P_j\langle g+1 \rangle[h-1]\) have the same level, any differential from a graded shift of \(P_i\) to a graded shift of \(P_j\) keeps the level unchanged.
  The other possibility is a map from \(P_i\langle g \rangle\) to \(P_i\langle g' \rangle\).
  Once again, it is clear from~\cref{rem:generating-morphisms} that such differentials have the form
  \[P_i \langle g \rangle[h] \to P_i\langle g' \rangle[h-1],\]
  where \(g'\) is either \(g\) or \(g + 2\).
  Any component of the form
  \[P_i \langle g \rangle[h] \to P_i\langle g \rangle[h-1]\]
  is necessarily a multiple of the identity map, and identity maps do not appear in the Khovanov--Seidel complex.
  Hence the only possibility is multiples of the loop map, which have the form
  \[P_i \langle g \rangle[h] \to P_i\langle g+2 \rangle[h-1].\]  
  Therefore any portion of the differential from a graded shift of a \(P_i\) to another graded shift of itself raises the level by one.

  Computing the \(j\)th cohomology of \(P^{\bullet}\) simply amounts to truncating the complex to level \(j\).
  The object \(P^{\bullet}\) is in the heart if and only if it is concentrated in a single cohomological degree, which means that its differential has no components of the form
  \[P_i \langle g \rangle[h] \to P_i\langle g+2 \rangle[h-1].\]
  This is the case if and only if the corresponding curve does not loop back around any point; that is, if and only if it is \(x\)-monotone.
  The lemma is proved.
\end{proof}

\subsection{Cohomology filtrations of spherical objects}
In this section we explain how wiggly pseudotriangulations govern the combinatorial structure of a natural categorical decomposition of spherical objects of \(\mathcal{C}_m\).

Recall from~\cref{defprop:cohomology-filtration} that any object \(X\) of \(\mathcal{C}_m\) has a canonical cohomology filtration whose factors are homological shifts of objects in \(\mathcal{A}_m\).
For general \(X\), the cohomology filtration can be arbitrary.
However, it turns out that the cohomology filtration is highly constrained if \(X\) is spherical, and these constraints are given by wiggly pseudotriangulations, as explained in the following proposition.
\begin{proposition}
  Let \(X \in \mathcal{C}_{n+1}\) be a spherical object.
  Consider its cohomology filtration
  \[
    \begin{tikzcd}[column sep=0.3cm]
      0 \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} & \cdots & H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]
  Consider the set of all nonzero objects \(\{H^j(X)\}\) taken without multiplicity, and let \(T\) be corresponding set of wiggly curves.
  We have the following results.
  \begin{enumerate}
  \item 
    \(T\) is a subset of a wiggly pseudotriangulation.
  \item \(T\) contains at most one of the two irrelevant arcs.
  \end{enumerate}
\end{proposition}
\begin{proof}
  The proof of this proposition is an easy consequence of analysing cohomology filtration from the point of view of arcs on the disk with marked points.
  \asilata{TODO.}
\end{proof}

\subsection{Categorical constraints on compatibility of curves}
In this section, we give a categorical criterion for two wiggly arcs to be compatible as in~\cref{def:compatible}.
As such, this section is somewhat dual to the previous section --- our criterion determines whether a pair of wiggly arcs can belong together in a single wiggly pseudotriangulation, whereas the previous section generated a wiggly pseudotriangulation given any object of the category.
\begin{proposition}
  Consider two wiggly arcs \(\alpha\) and \(\alpha'\), and let \(X\) and \(X'\) be objects in the standard heart \(\mathcal{A}_m\) corresponding to \(\alpha\) and \(\alpha'\) respectively.
  Then \(\alpha\) and \(\alpha'\) are compatible in the sense of~\cref{def:compatible} if and only if \(\Ext^1(X\langle g \rangle, X\langle g' \rangle) = 0\) for any two integers \(g\) and \(g'\).
\end{proposition}
\begin{remark}
  This proposition is similar to the condition of being compatible in a cluster-tilting object.
  \asilata{Expand and make precise.}
\end{remark}
\begin{remark}
  As stated, this proposition is not expected to be true in greater generality.
  However, we expect to see similar constraints on compatibility, which we hope to turn to in forthcoming work.
  \asilata{Decide what to say here, because the forthcoming work is far from complete, although I do have counterexamples in type \(D_4\). Discussing how to interpret these counterexamples may not involve too much more work?}
\end{remark}
\begin{question}
  How to best relate this section to the previous one purely categorically?
  That is, given the categorical constraint on compatibility of arcs, is it also clear that it gives a constraint on objects appearing in the cohomology filtration?
  \asilata{Make more precise.}
\end{question}
